---
title: "2023: Day 23"
date: 2023-12-23
author: 
  - name: Ella Kaye
categories: [grids, graphs, tidyverse, tidygraph, igraph, adventdrob, ⭐]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/23)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2023" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder.
```


```{r}
#| eval: !expr OK
library(aochelpers)
input <- aoc_input_data_frame(23, 2023) 
head(input)
```

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Find the longest path on a grid between given start and end points.
:::

```{r}
library(tidyverse)
library(adventdrob)
library(tidygraph)
library(igraph)

grid <- input |> 
	adventdrob::grid_tidy(X1) |> 
	mutate(name = row_number()) 

start_node <- grid |> 
	filter(row == 1 & value == ".") |> 
	pull(name)

end_node <- grid |> 
	filter(row == nrow(input) & value == ".") |> 
	pull(name)

grid_adj <- adventdrob::adjacent_join(grid, grid)

grid_adj_connected <- grid_adj |> 
	mutate(is_connected = case_when(
		value == "." & value2 == "." ~ TRUE,
		value == ">" & row2 == row & col2 == (col + 1) & value2 == "." ~ TRUE,
		value == "<" & row2 == row & col2 == (col - 1) & value2 == "." ~ TRUE,	
		value == "^" & row2 == (row - 1) & col2 == col & value2 == "." ~ TRUE,
		value == "v" & row2 == (row + 1) & col2 == col & value2 == "." ~ TRUE,
		value == "." & value2 == "v" & col2 == col & row2 == (row + 1) ~ TRUE,
		value == "." & value2 == "^" & col2 == col & row2 == (row - 1) ~ TRUE,
		value == "." & value2 == "<" & col2 == (col - 1) & row2 == row ~ TRUE,
		value == "." & value2 == ">" & col2 == (col + 1) & row2 == row ~ TRUE,
		.default = FALSE
	))

grid_nodes <- grid |> 
	select(name)

grid_edges <- grid_adj_connected |> 
	filter(is_connected) |> 
	select(from = name, to = name2)

forest_graph <- tbl_graph(nodes = grid_nodes, edges = grid_edges) 
all_paths_lengths <- all_simple_paths(forest_graph, start_node, end_node) |> 
	lengths() 

# need to subtract one as paths include start node,
# whereas we want number of steps, not number of nodes in the path
max(all_paths_lengths - 1)

```

The majority of the code today follows a pattern familiar from [Day 10](../10/index.qmd){target="_blank"} and [Day 21](../21/index.qmd){target="_blank"}. 
I'd used the [**tidygraph**](https://tidygraph.data-imaginist.com/index.html) package on both those days, but today it didn't have the functionality we need. 
With **tidygraph**, we can find the shortest path, but here we need the longest.
We're told that we can never step on the same node twice, 
which means that the path we're looking for is the longest *simple* path.
The [**igraph**](https://r.igraph.org) package has a function `all_simple_paths()`,
which returns a list, with one element per simple path between the given nodes.
Each element is a vector containing all nodes in that path. 
For this puzzle, we use `lengths()` to find the length of each of those vectors.
We then need to subtract one from each, as we need the number of steps, not number of nodes. 
The max of those gets us one gold star.



## Part 2

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
As Part 1, but with many more possible paths.
:::

I had thought (i.e. optimistically hoped) it would be straightforward to adapt the code to change the slopes to regular paths, then run the same code again, something like:

```{r}
#| eval: false
grid_no_slopes <- grid |> 
	mutate(value = if_else(value %in% c("<", ">", "^", "v"), ".", value)) 

grid_no_slopes_adj <- adventdrob::adjacent_join(grid_no_slopes, grid_no_slopes)

grid_no_slopes_adj_connected <- grid_no_slopes_adj |> 
	mutate(is_connected = if_else(value == "." & value2 == ".", TRUE, FALSE))

grid_no_slopes_edges <- grid_no_slopes_adj_connected |> 
	filter(is_connected) |> 
	select(from = name, to = name2)

forest_no_slopes_graph <- tbl_graph(nodes = grid_nodes, edges = grid_no_slopes_edges) 
all_paths_no_slopes_lengths <- all_simple_paths(forest_no_slopes_graph, start_node, end_node) |> 
	lengths() 

max(all_paths_no_slopes_lengths - 1)
```

That ran speedily on the example input, and gave the example answer.
However, on the full input, after the computer chugged away for a few minutes, 
I got the error `vector memory exhausted (limit reached?)`.

There must be a more efficient way, but that's beyond what I currently know how to do.
It's becoming increasingly clear to me that to be successful in the latter stages of Advent of Code, 
I'm going to need a much better understanding of data structures and algorithms.

##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>





