---
title: "2023: Day 14"
date: 2023-12-14
author: 
  - name: Ella Kaye
categories: [base R, matrices, ⭐]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/14)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2023" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder.
```


```{r}
#| eval: !expr OK
library(aochelpers)
input <- aoc_input_matrix(14, 2023)
head(input, c(6, 6))
```

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
'Tilt' a matrix to roll the `"O"`s until they reach a `"#"`.
:::

```{r}
# reverse order of rows
input_rev <- input[nrow(input):1, ]

# function to tilt one column and find the load
tilt <- function(col) {
	sorted <- col |> 
		factor(levels = c("#", ".", "O")) |> 
		split(cumsum(col == "#")) |> 
		lapply(sort) |> 
		do.call(c, args = _) 
	
	which(sorted == "O") |> 
		sum()
}

# apply to all columns and find total load
apply(input_rev, 2, tilt) |> 
	sum()                  
```

The puzzle talks about tilting north, 
but since the load is equal to the number of rows from the bottom,
in terms of associating the load with the row index, 
it's easier to reverse the order of the rows and think about tilting the matrix south.

Next, a function that will order one column.
The strategy is to split the column at the `"#"`s,
sort each element of the resulting list to move the `"O"`s (having converted the vector into a factor so we can sort in the desired order),  
then rejoin the vector. 
We can then find the indices of the round rocks, 
and their sum is the load for that column.

Finally, apply that to each column of the matrix,
and find the total sum.

## Part 2

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Something to do with cycles.
:::

This doesn't seem well-suited to a brute force approach.
At the time of writing, I'm not familiar with any algorithm that would be suitable here. 
On the R4DS Slack, Jonathan Carroll and Tan Ho have hinted about using the cycle-detection algorithm and have posted their solutions.
At some point, I may come back to this, 
read about the algorithm and either attempt to implement it myself, or work through their code to get a deeper understanding.


##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>





