---
title: "2022: Day 2"
date: 2022-12-2
author: 
  - name: Ella Kaye
categories: [base R, strings]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2022/day/2)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2022" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder
```


```{r}
#| eval: !expr OK
library(aochelpers)
input <- aoc_input_vector(2, 2022)
head(input)
```

The strategy here is to write a function that calculates the score for each individual round, 
then use `sapply()` to apply it to each round in the input, and finally sum the results.
The [script](script.R) has a few more details.

```{r}
my_score <- function(round) {
    
    draw <- c("A X", "B Y", "C Z")
    win <- c("A Y", "B Z", "C X")
    lose <- c("A Z", "B X", "C Y")
    
    my_shape <- strsplit(round, " ")[[1]][2]
    shape_score <- switch(my_shape, "X" = 1, "Y" = 2, "Z" = 3)
    
    if (round %in% draw) outcome_score <- 3
    else if (round %in% win) outcome_score <- 6
    else outcome_score <- 0
    
    shape_score + outcome_score
}

sapply(input, my_score) |> sum()
```

## Part 2

Very similar for part 2, expect that the interpretation of the rules is different,
so we need to modify the function for caluclating the score.

```{r}
my_new_score <- function(round) {
    
    pick_scissors <- c("A X", "B Z", "C Y")
    pick_rock <- c("A Y", "B X", "C Z")
    pick_paper <- c("A Z", "B Y", "C X")
    
    outcome <- strsplit(round, " ")[[1]][2]
    outcome_score <- switch(outcome, "X" = 0, "Y" = 3, "Z" = 6)
    
    if (round %in% pick_rock) shape_score <- 1
    else if (round %in% pick_paper) shape_score <- 2
    else shape_score <- 3
    
    shape_score + outcome_score
}

sapply(input, my_new_score) |> sum()
```


##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>




