---
title: "2023: Day 24"
date: 2023-12-24
author: 
  - name: Ella Kaye
categories: [base R, line equations, aochelpers, ⭐]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/24)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2023" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder.
```


```{r}
#| eval: !expr OK
library(aochelpers)
input <- aoc_input_vector(24, 2023)
head(input)
```

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Determine which pairs of lines intersect within a given area, in the future.
:::

```{r}
#| cache: true
# path is one line of the original input
get_line <- function(path) {
	nums <- aochelpers::extract_numbers(path)
	
	px <- nums[1]
	py <- nums[2]
	vx <- nums[4]
	vy <- nums[5]
	m <- vy/vx
	b <- py - m * px
	
	return(list(px = px, py = py, m = m, b = b, vx = vx, vy = vy))
}

both_in_range <- function(x, 
                          y, 
                          min = 200000000000000, 
                          max = 400000000000000) {
	ifelse(x >= min && x <= max && y >= min && y <= max,
				 TRUE,
				 FALSE)
} 

paths_intersect <- function(path1, path2) {
	A <- get_line(path1)
	B <- get_line(path2)
	
	# parallel lines won't intersect
	if (A$m == B$m) {
		# print("lines are parallel")
		return(FALSE)
	}
	
	# point of intersection
	x <- (B$b - A$b)/(A$m - B$m)
	y <- A$m * x + A$b
	
	# check if in past
	if (x > A$px && A$vx < 0 || x < A$px && A$vx > 0) {
		#print("intersect in past for A")
		return(FALSE)
	}
	
	if (x > B$px && B$vx < 0 || x < B$px && B$vx > 0) {
		#print("intersect in past for B")
		return(FALSE)
	}
	
	both_in_range(x, y)
}

# check all pairs
# need stringsAsFactors = FALSE to enable comparison to get unique pairs
all_pairs <- expand.grid(input, input, stringsAsFactors = FALSE)
unique_pairs <- all_pairs[all_pairs$Var1 < all_pairs$Var2, ]

mapply(paths_intersect, unique_pairs$Var1, unique_pairs$Var2) |> 
	sum()
```

Now that there have been a few puzzles in which we need to extract all the numbers from a string, I've added a function, `extract_numbers()` to my [**aochelpers**](https://ellakaye.github.io/aochelpers) package to do that,
which comes in handy today.

A bit of maths gets us the equation of a line and, for any pair of lines, their point of intersection. Then, it's just a case of checking whether they intersected in the past (the part of the puzzle that caught me out for a bit) and, if not, whether they intersect in the test area.

Finally we use `expand.grid()` to get all pairs of lines, then filter down to unique pairs.
We can then apply our `paths_intersect()` function to all unique pairs, using `mapply()` since `paths_intersect()` takes more than one argument. Since `TRUE` has a numeric value of `1` and `FALSE` is `0`, the sum of the output of `paths_intersect()` on all pairs gives the number of pairs of lines which intersect, as required.

## Part 2

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Find the starting position and velocity of a rock which,
when thrown, will hit all the hailstones.
:::

Not attempted (yet).

##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>





