---
title: "2023: Day 25"
date: 2023-12-25
author: 
  - name: Ella Kaye
categories: [tidyverse, graphs, igraph, ⭐]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/25)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2023" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder.
```


```{r}
#| eval: !expr OK
library(aochelpers)
input <- aoc_input_vector(25, 2023)
head(input)
```

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Given a graph, remove three edges to split the graph into two separate disconnected groups, then multiply the sizes of those groups together.
:::

Learning graph theory at 6am -- that's quite a start to a Christmas day!

At first, I had no idea how to approach this problem and figured there must be some graph theory that would achieve exactly what was required, 
preferably with an algorithm already implemented in the [**igraph**](https://r.igraph.org) package. 
A bit of googling and I was in luck: we need a [minimum cut](https://en.wikipedia.org/wiki/Minimum_cut). 
In this case, the minimum number of edges that can be removed to split the graph into two groups.^[The graph-theoretic term for these groups is *components* but presumably that's not being used so as not to cause confusion with the machine components in the puzzle text.] 
This is implemented in **igraph** with the `min_cut()` function,
which can include in its output the nodes in each group. 
We then multiply the lengths of these vectors together for the puzzle answer. 
Before that, the new family of `separate_*` functions in **tidyr** makes it easy to wrangle the input into a two-column data frame representing edges, which we turn to a matrix to make it acceptable for `matrix_from_edgelist()`. 
Putting it together, the solution is concise and very fast, 
which is just as well because it's Christmas Day and I don't have a lot of time for coding!

```{r}
library(tidyverse)
library(igraph)

input_tbl <- input |> 
	as_tibble() |> 
	separate_wider_delim(value, ":", names = c("from", "to")) |> 
	separate_longer_delim(to, " ") |> 
	filter(to != "")

edgelist <- as.matrix(input_tbl)
comp_graph <- graph_from_edgelist(edgelist, directed = FALSE)
groups <- min_cut(comp_graph, value.only = FALSE)
length(groups$partition1) * length(groups$partition2)
```


## Part 2

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Have solved all other problems!
:::

So much for not much coding required to complete Day 25! 
There's no additional puzzle for the final star -- we "only" need to have acquired the previous 49 stars.

As of Christmas Day 2023, I have 30 gold stars -- 19 to go! 
That's a task for another day.

Happy holidays!

##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>





