---
title: "2023: Day 5"
date: 2023-12-5
author: 
  - name: Ella Kaye
categories: [base R, tidyverse, loops, scalability issues, ⭐]
draft: false
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/5)

[My data](input){target="_blank"}

## Part 1

```{r}
#| echo: false
OK <- "2023" < 3000
# Will only evaluate next code block if an actual year has been substituted for the placeholder.
```


```{r}
#| eval: !expr OK
library(aochelpers)
library(tidyverse)
#input <- aoc_input_vector(5, 2023)
input <- read_lines(here::here("2023", "day", "5", "example-input"))
head(input)
```

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Given a set of seeds, and a set of maps, use the maps to find the location of the closest seed.
:::

Big struggle today. 
On December 5th itself, I finally produced a solution that worked on the example input,
even though I knew by then that my approach wouldn't work on actual input.
My first attempt was a strange mash-up of base R and tidyverse, via various digressions.
It runs into memory issues on the full input. 
I was so determined at least to get something on the example input that I ploughed on with it regardless,
even introducing a fudge to make it work on the specific example.

I'll document the wrong approach here, as a reminder of what not to do, 
then show a working solution that I figured out the following day, 
thanks to a tip from Tan Ho on the R4DS Slack.

The first part, wrangling the data, is the same for both approaches.

Wrangle the data:

```{r}
# Split the input at the blank lines, then remove the blank lines
split_input <- split(input, cumsum(input == "")) |> 
	lapply(\(x) x[x != ""])

# Get a vector of the seeds
seeds <- split_input[[1]] |> 
	aochelpers::extract_numbers()

# A list of the maps, without the map names
maps <- tail(split_input, -1) |> 
	lapply(\(x) tail(x, -1))
```

### A wrong approach

For each map, make a tibble for every possible source-destination pair. 
In retrospect, this is where I started to go wrong.
These tibbles are billions of lines long on the full input, 
and we don't need every possible pair.

Take a single item from a map, e.g. `"50 98 2"` and make a tibble:

```{r}
make_map <- function(x) {
	n <- x[3]
	tibble(
		source = x[2]:(x[2]+n-1),
		destination = x[1]:(x[1]+n-1)
	)
}
```

Take a whole map, use `make_map` for each map item, and bind into a big data frame.
This is where the it fails on the full input. 
The combined data frame runs out of memory.
By the time I put in the fudge to make this work on the example input, 
I knew that my approach wouldn't work on the full input, which is why it's only part implemented.
I would have needed a better approach for tracking source elements that aren't covered by the map.
```{r}
make_map_df <- function(map, vec = NULL) {
	
	map <- map |> 
		lapply(extract_numbers) |>
		lapply(make_map) |>
		bind_rows() 
	
	# account for elements in source not in a range on the map
	# only implement for elements less than the minimum source
	# as a fudge to get this to work on the example data.
	if (!is.null(vec)) {
		
		min_source <- min(map$source)
		min_vec <- min(vec)
		
		if (min_vec < min_source) {
			extra_vec <- min_vec:(min_source-1)
			extra_rows <- tibble(source = extra_vec, destination = extra_vec) 
			map <- bind_rows(extra_rows, map)
		}
	} 
	
	map |> 
		arrange(source)
}
```

A function that joins two map tables to link the destination with the source.
This adds even more rows! 
It did, at least, give me an opportunity to learn about the new `join_by()` function in dplyr v1.1.

```{r}
build_map_table <-  function(map_table, map) {
	full_join(map_table, map, join_by(destination == source)) |> 
		mutate(destination.y = if_else(is.na(destination.y), destination, destination.y)) |> 
		select(-destination) |> 
		rename(destination = destination.y)
}
```

Start with the seed-to-soil map, then iterate over the maps, 
joining them to make new source-to-destination links.

```{r}
map_table <- make_map_df(maps[[1]], seeds) 
for (i in 2:length(maps)) {
	map_df <- make_map_df(maps[[i]])
	map_table <- build_map_table(map_table, map_df)
}
```

In the final `map_table`, the source column represents the seeds 
and the destination column represents the location. 
We have our minumum location, at least for the example data.

```{r}
map_table |> 
	filter(source %in% seeds) |> 
	summarise(min_location = min(destination)) |>
	pull(min_location)
```

Lessons learnt:

- spend time looking at actual input and thinking of an appropriate strategy for that scale
- don't plough on with a solution for the example input when I know it won't work on the full input

### A right approach

A tip from Tan Ho to work with information describing the the ranges, 
rather than expanding them in full, got me on the right track.
I also wasn't helped above by the mash-up of tidyverse and base R approaches,
so I went back to the familiar strategy of writing a function that takes one seed and returns its location, 
and applying it to all seeds to find the minimum.
We need a few helper functions first.

```{r}
# `source` is a single element in the source, e.g. seeds[1], 
# and `item` is a single line of a map, e.g. "50 98 2"
# returns a logical indicating whether the source is in the range 
# described by the item
source_in_range <- function(source, item) {
	item <- aochelpers::extract_numbers(item)
	
	if (source >= item[2] && source <= item[2] + item[3] - 1) {
		return(TRUE)
	}
	else {
		return(FALSE)
	}
}
```

```{r}
# source is as above
# map is a full map, i.e. a single elements of `maps`
# returns a single element, the value of the destination
source_to_destination <- function(source, map) {
	
	# check if source is in each range
	# logical vector of length(map)
	is_in_range <- purrr::map_lgl(map, \(x) source_in_range(source, x))
	which_range <- which(is_in_range)
	
	if (length(which_range) == 0) {
		return(source)
	}
	else {
		# get the item that contains the source
		item <- map[which_range] |> extract_numbers()
		destination <- source - item[2] + item[1]
		return(destination)
	}
}
```

```{r}
# starting with a seed, this function iterates over all the maps 
# to return the value of the location.
seed_to_location <- function(seed, maps) {
	
	# start with the seed
	source <- seed
	
	# iterate over maps
	for (i in seq_along(maps)) {
		# get the destination
		destination <- source_to_destination(source, maps[[i]])
		# set the source to the destination
		source <- destination
		#cat(paste0("round ", i, ": source ", source, "\n"))
	}
	
	# return the final destination
	destination
}
```

Now let's read in our actual input and wrangle it as before:

```{r}
input <- aoc_input_vector(5, 2023)
split_input <- split(input, cumsum(input == "")) |> 
	lapply(\(x) x[x != ""])

seeds <- split_input[[1]] |> extract_numbers()

maps <- tail(split_input, -1) |> 
	lapply(\(x) tail(x, -1))
```

Finally, we apply `source_to_destination()` to all the seeds and find the minimum:

```{r}
seeds |> 
	sapply(\(x) seed_to_location(x, maps)) |> 
	min()
```

Got there!

## Part 2

::: {.callout-note collapse="false" icon="false"}
## The crux of the puzzle
Find the lowest location number for any initial seed, 
but from a *much* larger set of seeds than in Part 1.
:::

Not attempted (yet).


##### Session info {.appendix}

<details><summary>Toggle</summary>

```{r}
#| echo: false
library(sessioninfo)
# save the session info as an object
pkg_session <- session_info(pkgs = "attached")

# get the quarto version
quarto_version <- system("quarto --version", intern = TRUE)

# inject the quarto info
pkg_session$platform$quarto <- paste(
  system("quarto --version", intern = TRUE), 
  "@", 
  quarto::quarto_path()
  )

# print it out
pkg_session
```

</details>





